Fusing Filters with Integer Linear Programming

### 2014-06-20 2014betterfusionforfilters/section/03-SizeInference.tex
change function to program in size inference. conclusion changes.

% -----------------------------------------------------------------------------
\subsection{Constraint Generation}
The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}. The top level judgment ~~$\SizeF{\program}{\sigma}$~~ assigns a size scheme to a function. It does this by extracting size constraints and then solving them. This rule, along with the constraint solving process is discussed in the next section. The judgment ~~$\SizeB{\Gamma_1}{zs}{b}{\Gamma_2}{C}$ reads: ``Under environment~$\Gamma_1$, array variable $zs$ binds the result of $b$, producing a result environment $\Gamma_2$ and size constraints $C$''. The remaining judgment that extracts constraints from a list of bindings is similar. The environment $\Gamma$ has the following grammar:
$$
\Gamma~ @::=@ ~~\cdot ~~|~~ \Gamma,~ \Gamma ~~|~~ zs : k ~~|~~ k ~~|~~ \exists k
$$

As usual, $\cdot$ represents the empty environment and  $\Gamma,~ \Gamma$
environment concatenation. The element $zs : k$ records the size $k$ of some
array variable $zs$. A plain $k$ indicates that $k$ can be unified with other
size types when solving constraints, whereas $\exists k$ indicates a  \emph{rigid} size variable that cannot be unified with other sizes. We use the $\exists k$ syntax because this variable will also be existentially quantified if it appears in the size scheme of the overall function.

@@

For example, the final environment and constraints generated for the @normalize2@ example from Section~\ref{s:Introduction} are as follows:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & \true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  \true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$
Rule~(SFun) also characterises the functions we accept: A function is \emph{valid} if and only if $\exists \sigma.\ \SizeF{\program}{\sigma}$.

% -------------------------------------------------------------------
\subsection{Constraint Solving and Generalization}
The top-level rule in Figure~\ref{f:ConstraintGeneration} assigns a size scheme to a function by first extracting size constraints, before solving them and generalizing the result. In the rule, the solving process is indicated by $\textrm{SOLVE}$, and takes an environment and a constraint set, and produces a solved environment and constraint set. As the constraint solving process is both standard and straightforward, we only describe it informally.

@@

To produce the final size scheme, we look up the sizes of the input and output variables of the original function from the solved constraints and generalize appropriately. This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}. In this case, no rigid size variables appear in the result, so we can universally quantify all size variables.
$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

@@

% -----------------------------------------------------------------------------
\subsection{Rigid Sizes}
When the environment of our size constraints contains rigid variables (indicated by $\exists k$), we introduce existential quantifiers instead of universal quantifiers into the size scheme. Consider the @filterLeft@ function from Section~\ref{s:SizeTypes}
\begin{code}
      filterLeft xs
        = let ys1 = map (+ 1)   xs
              ys2 = filter even xs
          in (ys1, ys2)
\end{code}
The size constraints for this function, already in solved form, are as follows.
$$
\begin{array}{ll}
       & xs : k_{xs},~ ys_1 : k_{ys1},~ \exists k_1,~ ys_2 : k_{ys2},~ k_2
\\
\vdash &          k_{ys_1} = k_1
        ~\wedge~  k_{ys_2} = k_2
        ~\wedge~  k_{xs}   = k_2
\end{array}
$$

@@

Note that, although Rule~(SFun) from Figure~\ref{f:ConstraintGeneration} performs a \emph{generalisation} process, there is no corresponding instantiation rule. The size inference process works on the entire graph at a time, and there is no mechanism for one operator to invoke another. To say this another way, all subgraphs are fully inlined. Recall from \S\ref{s:CombinatorNormalForm}, that we assume our operator graphs are embedded in a larger host program. We use size information to guide the clustering process, and although the host program can certainly call the operator graph, static size information does not flow across this boundary.

When producing size schemes, we do not permit the arguments of an operator graph to have existentially quantified sizes. This restriction is necessary to reject programs that we cannot statically guarantee will be well sized. For example:
\begin{code}
    bad1 xs  = let flt   = filter p xs
                   ys    = map2   f flt xs
               in  ys
\end{code}
The above function filters its input array, and then applies @map2@ to the filtered version as well as the original array. As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true. The size constraints are as follows:
$$
\begin{array}{ll}
       & xs : k_{xs},~ flt : k_{flt},~ \exists k_1,~ ys : k_{ys},~ k_2
\\
\vdash &          k_{flt}  = k_1
        ~\wedge~  k_{flt}  = k_2
        ~\wedge~  k_{xs}   = k_2
        ~\wedge~  k_{ys}   = k_2
\end{array}
$$

@@

In this case, Rule~(SFun) does not apply, because the parameter variable $xs$ has size $k_1$, but $k_1$ is marked as rigid in the environment (with $\exists k_1$).

As a final example, the following function is ill-sized, because the two filter operators are not guaranteed to produce the same number of elements.
\begin{code}
     bad2 xs = let as  = filter p1 xs
                   bs  = filter p2 xs
                   ys  = map2   f  as bs
               in  ys
\end{code}

The initial size constraints for this function are:
$$
\begin{array}{ll}
       & xs : k_{xs},~ as : k_{as},~ \exists k_1,~ bs : k_{bs},~ \exists k_2,~ ys : k_{ys},~ k_3
\\
\vdash &          k_{as}   = k_1
        ~\wedge~  k_{bs}   = k_2
        ~\wedge~  k_{as}   = k_3
        ~\wedge~  k_{bs}   = k_3
        ~\wedge~  k_{ys}   = k_3
\end{array}
$$

@@

\textbf{Lemma: sole transducers}.
If a function $f$ is \emph{valid}, then its bindings will have at most one transducer:
\[
\forall f, \sigma, n.\ f :_s \sigma \implies |trans(binds(f), n)| \le 1
\]
% \textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
% then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
% As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
% if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
% The other cases are trivial.

@@

\textbf{Lemma: sole parents}.
% I replaced 'one set of parents' with 'one pair of parents' here, because that's what the lemma says
For some bindings $f$ with valid constraints, each pair of names $a$ and $b$ will have at most one pair of parents $parents(a,b)$.
\[
\forall f, \sigma, a, b.\ f :_s \sigma \implies |parents(binds(f), a, b)| \le 1
\]
% \textbf{Proof:} since each binding only has at most one transducer, and transducers' iteration sizes are not the same, there will be at most one pair of transducers up the chain that have equal types.

These two lemmas are used in the integer linear programming formulation, when constructing the program.
When fusing two nodes of different iteration size, at most one pair of parents will need to be checked.

>>>>

% -----------------------------------------------------------------------------
\subsection{Constraint Generation}
The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}. The top level judgment ~~$\SizeF{\program}{\sigma}$~~ assigns a size scheme to a program. It does this by extracting size constraints and then solving them. This rule, along with the constraint solving process is discussed in the next section. The judgment ~~$\SizeB{\Gamma_1}{zs}{b}{\Gamma_2}{C}$ reads: ``Under environment~$\Gamma_1$, array variable $zs$ binds the result of $b$, producing a result environment $\Gamma_2$ and size constraints $C$''. The remaining judgment that extracts constraints from a list of bindings is similar. The environment $\Gamma$ has the following grammar:
$$
\Gamma~ @::=@ ~~\cdot ~~|~~ \Gamma,~ \Gamma ~~|~~ zs : k ~~|~~ k ~~|~~ \exists k
$$

As usual, $\cdot$ represents the empty environment and  $\Gamma,~ \Gamma$
environment concatenation. The element $zs : k$ records the size $k$ of some
array variable $zs$. A plain $k$ indicates that $k$ can be unified with other
size types when solving constraints, whereas $\exists k$ indicates a  \emph{rigid} size variable that cannot be unified with other sizes. We use the $\exists k$ syntax because this variable will also be existentially quantified if it appears in the size scheme of the overall program.

@@

For example, the final environment and constraints generated for the @normalize2@ example from Section~\ref{s:Introduction} are as follows:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & \true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  \true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$
Rule~(SProgram) also characterises the programs we accept: a program is \emph{valid} if and only if $\exists \sigma.\ \SizeF{\program}{\sigma}$.

% -------------------------------------------------------------------
\subsection{Constraint Solving and Generalization}
The top-level rule in Figure~\ref{f:ConstraintGeneration} assigns a size scheme to a program by first extracting size constraints, before solving them and generalizing the result. In the rule, the solving process is indicated by $\textrm{SOLVE}$, and takes an environment and a constraint set, and produces a solved environment and constraint set. As the constraint solving process is both standard and straightforward, we only describe it informally.

@@

To produce the final size scheme, we look up the sizes of the input and output variables of the original program from the solved constraints and generalize appropriately. This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}. In this case, no rigid size variables appear in the result, so we can universally quantify all size variables.
$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

@@

% -----------------------------------------------------------------------------
\subsection{Rigid Sizes}
When the environment of our size constraints contains rigid variables (indicated by $\exists k$), we introduce existential quantifiers instead of universal quantifiers into the size scheme. Consider the @filterLeft@ program from Section~\ref{s:SizeTypes}
\begin{code}
      filterLeft xs
        = let ys1 = map (+ 1)   xs
              ys2 = filter even xs
          in (ys1, ys2)
\end{code}
The size constraints for this program, already in solved form, are as follows.
$$
\begin{array}{ll}
       & xs : k_{xs},~ ys_1 : k_{ys1},~ \exists k_1,~ ys_2 : k_{ys2},~ k_2
\\
\vdash &          k_{ys_1} = k_1
        ~\wedge~  k_{ys_2} = k_2
        ~\wedge~  k_{xs}   = k_2
\end{array}
$$

@@

Note that, although Rule~(SProgram) from Figure~\ref{f:ConstraintGeneration} performs a \emph{generalisation} process, there is no corresponding instantiation rule. The size inference process works on the entire graph at a time, and there is no mechanism for one operator to invoke another. To say this another way, all subgraphs are fully inlined. Recall from \S\ref{s:CombinatorNormalForm}, that we assume our operator graphs are embedded in a larger host program. We use size information to guide the clustering process, and although the host program can certainly call the operator graph, static size information does not flow across this boundary.

When producing size schemes, we do not permit the arguments of an operator graph to have existentially quantified sizes. This restriction is necessary to reject programs that we cannot statically guarantee will be well sized. For example:
\begin{code}
    bad1 xs  = let flt   = filter p xs
                   ys    = map2   f flt xs
               in  ys
\end{code}
The above program filters its input array, and then applies @map2@ to the filtered version as well as the original array. As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true. The size constraints are as follows:
$$
\begin{array}{ll}
       & xs : k_{xs},~ flt : k_{flt},~ \exists k_1,~ ys : k_{ys},~ k_2
\\
\vdash &          k_{flt}  = k_1
        ~\wedge~  k_{flt}  = k_2
        ~\wedge~  k_{xs}   = k_2
        ~\wedge~  k_{ys}   = k_2
\end{array}
$$

@@

In this case, Rule~(SProgram) does not apply, because the parameter variable $xs$ has size $k_1$, but $k_1$ is marked as rigid in the environment (with $\exists k_1$).

As a final example, the following program is ill-sized, because the two filter operators are not guaranteed to produce the same number of elements.
\begin{code}
     bad2 xs = let as  = filter p1 xs
                   bs  = filter p2 xs
                   ys  = map2   f  as bs
               in  ys
\end{code}

The initial size constraints for this program are:
$$
\begin{array}{ll}
       & xs : k_{xs},~ as : k_{as},~ \exists k_1,~ bs : k_{bs},~ \exists k_2,~ ys : k_{ys},~ k_3
\\
\vdash &          k_{as}   = k_1
        ~\wedge~  k_{bs}   = k_2
        ~\wedge~  k_{as}   = k_3
        ~\wedge~  k_{bs}   = k_3
        ~\wedge~  k_{ys}   = k_3
\end{array}
$$

@@

\textbf{Lemma: sole transducers}.
If a program $p$ is \emph{valid}, then its bindings will have at most one transducer:
\[
\forall p, \sigma, n.\ p :_s \sigma \implies |trans(binds(p), n)| \le 1
\]
% \textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
% then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
% As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
% if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
% The other cases are trivial.

@@

\textbf{Lemma: sole parents}.
% I replaced 'one set of parents' with 'one pair of parents' here, because that's what the lemma says
For some program $p$ with valid constraints, each pair of names $a$ and $b$ will have at most one pair of parents $parents(a,b)$.
\[
\forall p, \sigma, a, b.\ p :_s \sigma \implies |parents(binds(p), a, b)| \le 1
\]
% \textbf{Proof:} since each binding only has at most one transducer, and transducers' iteration sizes are not the same, there will be at most one pair of transducers up the chain that have equal types.

These two lemmas are used in the integer linear programming formulation, when generating the constraints.
When fusing two nodes of different iteration size, at most one pair of parents will need to be checked.

### 2014-06-20 2014betterfusionforfilters/section/07-Conclusion.tex
change function to program in size inference. conclusion changes.

For this work to be useful for Data Parallel Haskell, more combinators are required such as segmented fold, segmented map, which operate on segmented representations of nested arrays.
Rate inference will have to be adapted to handle segmented arrays, possibly using an inner and outer rates.
For example, segmented fold takes a segmented array, and returns a single array of the outer rate.

>>>>

For this work to be useful for Data Parallel Haskell, more combinators are required such as segmented fold and segmented map, which operate on segmented representations of nested arrays.
Rate inference will have to be adapted to handle segmented arrays, possibly using an inner and outer rates.
For example, segmented fold takes a segmented array, and returns a single array of the outer rate.

Data flow fusion, which generates the sequential loops, must also be extended with these extra combinators.
Data flow fusion can also be extended to generate parallel loops for most combinators by simply splitting the workload among processors.
Merging the output of each loop can be performed by concatenating the results of filters, and merging the folds, assuming the fold operation is associative.

Another interesting possibility is combinators where the output order is not important.
There are two orderings for cross product: one that requires the second array to be manifest, the other requires the first array to be manifest.
It may be worthwhile to add a separate cross combinator that does not ensure which ordering is used, but instead uses the ordering that produces the best clustering.
This could be achieved in the integer linear program by requiring that at least one of the cross combinator's inputs are not fused together.
